---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r load_packages}
library(dplyr)
library(ggplot2)
myactivity<-read.csv("activity.csv",stringsAsFactors = FALSE,na.strings = "NA")
myactivity$date<-as.Date(myactivity$date,"%Y-%m-%d")
```

## What is the total number of steps taken per day?

```{r calculate_total}
myactivity_bydate<-group_by(myactivity,date)
mysum<-summarise(myactivity_bydate,total_steps_by_day=sum(steps))
mysum<-as.data.frame(mysum)
hist(mysum$total_steps_by_day,ylim=c(0,35),xlab="total number of steps per day",main="Historgram of Total Number of Steps per Day")
```


## What is the average daily activity pattern?
###calculate and plot mean steps per day
```{r calculate_mean}
mymean<-summarise(myactivity_bydate,mean_steps_by_day=mean(steps,na.rm = TRUE))
mymean<-as.data.frame(mymean)
plot(mymean$date,mymean$mean_steps_by_day,type="l",xlab="Date",ylab="Mean Number of Steps per Day",main="Mean Number of Steps per Day - Some Missing Data")
#alternatively use ggplot
#g1<-ggplot(mymean,aes(x=date,y=mean_steps_by_day))
#g1+geom_line()+labs(title="Mean Number of Steps per Day - Some Missing Data")+
#        labs(x="Date",y="Mean Number of Steps per Day")
```

###
###calculate median per day
```{r calculate_median}
mymedian<-summarise(myactivity_bydate,median_steps_by_day=median(steps,na.rm = TRUE))
mymedian<-as.data.frame(mymedian)
```

## Imputing missing values by replacing them with means of that particular internval
###calcualte mean per interval
```{r calculate_interval_mean}
myactivity_byinterval<-group_by(myactivity,interval)
mymean_interval<-summarise(myactivity_byinterval,mean_by_interval=mean(steps,na.rm = TRUE))
mymean_interval<-as.data.frame(mymean_interval)
```

###now fill in missing steps in original dataset, replace by mean of that interval

```{r fill_missing_data}
temp1<-merge(myactivity,mymean_interval)
temp1<-arrange(temp1,date,interval)
for (i in 1:nrow(temp1)){
        if (is.na(temp1[i,]$steps)==TRUE) {
                temp1[i,]$steps <-temp1[i,]$mean_by_interval
        }
}
myactivity_nomissingdata<-temp1[,1:3]
```
###now recalculate sum, mean and median for each day
```{r recalculate_without_missingdata}
myactivity_nomissingdata_bydate<-group_by(myactivity_nomissingdata,date)
mysum2<-summarise(myactivity_nomissingdata_bydate,total_steps_by_day=sum(steps))
mysum2<-as.data.frame(mysum2)
hist(mysum2$total_steps_by_day,ylim=c(0,35), xlab="total number of steps per day",main="Historgram of Total Number of Steps per Day - No Misssing Data")

mymean2<-summarise(myactivity_nomissingdata_bydate,mean_steps_by_day=mean(steps,na.rm = TRUE))
mymean2<-as.data.frame(mymean2)
plot(mymean2$date,mymean2$mean_steps_by_day,type="l",xlab="Date",ylab="Mean Number of Steps per Day",main="Mean Number of Steps per Day - No Missing Data")

mymedian2<-summarise(myactivity_nomissingdata_bydate,median_steps_by_day=median(steps,na.rm = TRUE))
mymedian2<-as.data.frame(mymedian2)
```

###now plotting Side by Side the sum and mean of steps per day before and after data imputation
```{r plot_all}
par(mfrow=c(2,2))
hist(mysum$total_steps_by_day,ylim=c(0,35),xlab="total number of steps per day")
hist(mysum2$total_steps_by_day,ylim=c(0,35), xlab="total number of steps per day")
plot(mymean$date,mymean$mean_steps_by_day,type="l",xlab="Date",ylab="Mean Number of Steps per Day")
plot(mymean2$date,mymean2$mean_steps_by_day,type="l",xlab="Date",ylab="Mean Number of Steps per Day")
```

## Are there differences in activity patterns between weekdays and weekends?
### add factor weekday column 
```{r add_weekday_factor}
temp2<-mutate(myactivity_nomissingdata,whichday=weekdays(date))
temp2<-mutate(temp2,weekday=whichday)
for (i in 1:nrow(temp2)){
        if (temp2[i,]$whichday %in% c("Saturday","Sunday") == TRUE) {
                temp2[i,]$weekday <- "weekend"
        }
        else {
                temp2[i,]$weekday <- "weekday"
        }
}
temp2$weekday<-factor(temp2$weekday)
```
### now plot mean number of steps per interval for weekday and weekends*
```{r calculate_plot_weekday_weekend_steps}
temp3<-group_by(temp2,interval,weekday)
temp4<-summarise(temp3,mean_by_interval=mean(steps))
temp4<-as.data.frame(temp4)
qplot(interval,mean_by_interval,data=temp4,ylim=c(0,250),ylab="Mean number of steps",facets = .~weekday) + geom_line()+ geom_smooth(method = "lm", se = FALSE)
#now try ggplot2
g2<-ggplot(temp4,aes(x=interval,y=mean_by_interval))
g2+geom_line()+facet_grid(.~weekday)+geom_smooth(method = "lm", se = FALSE)+labs(y="Mean Number of Steps")

```



###It can be observed from these graphs, weekends are more active after mid-late morning around 10am;
###While weekdays are more active in the early-mid morning period before 10am